---
title: "statCanR"
author: "Thierry Warin"
data: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{statCanR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(statcanR)
```

# Goals 

This vignette describes the code architecture of the _statcanR_  R package. This package allows to get all Statistics Canada' open economic data (formerly known as CANSIM tables) now identified by product IDs (PID) by the new Statistics Canada's Web Data Service.

In details, the _sqs_statcan_data()_ function has 2 arguments:   
  
  1. _tableNumber_
  2. _lang_
  
  
The _tableNumber_ argument simply referes to the table number of the Statistics Canada data table a user wants to collect, such as '_14-10-0287-03_' for the _Labour force characteristics by province, monthly, seasonally adjusted_ as an example.

The second argument, _lang_, referes to the language. As Canada is a bilingual country, Statistics Canada displays all the economic Data in both languages. Therefore, users can choose to collect satistics data tables in French or English by setting the _lang_ argument with c("fra", "eng").

The code architecture of the _sqs_statcan_data()_ function is the following one:

The first step if to clean the table number in order to align it with the official typology of Statistic Canada's Web Data Service. The second step is to create a temporary folder where all the data wrangling operations are performed. The third step is to check and select the correct language. The fourth step is to define the right URL where the actual data table is stored and  then to download the _.Zip_ container. The fifth step is to unzip the previously downloaded _.Zip_ file to extract the data and metadata  from the _.csv_ files. The final step is to load the statistics data into a data frame called '_sqs_data_' and to add the official table indicator name in the new '_INDICATOR_' column.         
  

To be more precise about the _sqs_statcan_data()_ function, here is below a further ode description:   
  
  - `tableNumber <- gsub("-", "", substr(tableNumber, 1, nchar(tableNumber)-2))`:
  The first step is to clean the table number provided by the user in order to collect the overall data table related to the specific indicator the user is interested in. In fact, each indicator is an excerpt of the overall table.
  In addition, following the new Statistics Canada's Web Data Service, the URL typology defined by REST API is stored in csv files by table numbers without a '-'. Also, the last 2 digits after the last '-' refer to the specific excerpt of the original table.
  Therefore, following the Statistics Canada Web Data Service's typology, the function first removes the '-' and the last 2 digits from the user's selection.

  - `if(lang == "eng")` | `if(lang == "fra")`:
  The second step is the '_if statement_' to get the data in the correct language.

  - `urlFra <- paste0("https://www150.statcan.gc.ca/n1/fr/tbl/csv/", tableNumber, "-fra.zip")`:
  The third step is to create the correct url in order to download the respective _.Zip_ file from the Statistics Canada Web Data Service.   
  
  - `download.file(urlEng, destfile=paste0(path,"/temp/datasetEng.zip"), method="curl")`:
  The fourth step is a simple downloading function that extracts _.Zip_ file and download it into a temporary folder.
  
  - `unzip(paste0(path, "/temp/datasetEng.zip"), exdir = paste0(path,"/temp"))`:
  The fifth step consists in unzipping the _.Zip_ file. The unzipping process gives access to two diffrent _.csv_ files, such as the overall data table and the metadata table.
    
  - `data <- fread(paste0(path,"/temp/", tableNumber, ".csv"))`:
  The sixth step consists in loading the data table into a unique data frame. The _fread_ function from the _data.table_ package is used for its higher performance.   
    
  - `data$INDICATOR <- as.character(0)` and `data$INDICATOR <- as.character(read.csv(paste0(path,"/temp/", tableNumber, "_MetaData.csv"))[1,1])`:
  The seventh step of the _sqs_statcan_data()_ function consists in adding the name of the table from the metadata table. 
  
  - `unlink(paste0(path,"/temp/"), recursive = TRUE)`:
  The eighth step deletes the temporary folder used to download, unzip, extract and load the data.
  
  - `return(data)`:
  Finally, the last step of the _sqs_statcan_data()_ function allows to return the value into the user's environment.



# Practical usage

This section presents an example of how to use the _statcanR_ R package and its function _sqs_statcan_data()_.    
  

The following example is provided to illustrate how to use the function. It consists in collecting some descriptive statistics about the Canadian Labour Force at the federal, provincial and industrial levels, on a monthly basis.  

With a simple web search '_statistics canada wages by industry metropolitan area monthly_', the table number can easily be found on Statisitcs Canada's webpage. Here is below a figure that illustrates this example, such as '_14-10-0063-01_' for the _Employee wages by industry, monthly, unadjusted for seasonality at the ferederal and provincial levels_.

Once the table number is identified, the _sqs_statcan_data()_ function is easy to use in order to collect the data, as following:  
  
```{r, eval=FALSE}
# Get data with sqs_statcan_data function
mydata <- sqs_statcan_data("14-10-0063-01", "eng")
```
